---
title: "Untitled"
author: "Rayan Slatni"
date: "2023-07-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, warning=FALSE}
PACKAGES <- c(
  "caret",          # To create data partitions
  "genefilter",     # To remove low count genes
  "ggplot2",        # To plot and save images
  "gplots",
  "ggpubr",
  "glmnet",         # To perform lasso-ridge
  "MASS",           # To use LDA function 
  "pROC",           # To plot ROC curves
  "randomForest",   # To use random forest algorithm
  "RColorBrewer",   # To define heatmap palette
  "reshape2",       # To melt dataset for plotting
  "ROCR",           # To plot ROC curves 
  "tidyverse",      # To make data handling easier
  "GEOquery" ,       #for access to the databse
  "useful",
  "factoextra",  # To get better plots (ggplot wrapper)
  "ggfortify",      #To plot PCA
  "vctrs",
  "tibble",
  "e1071",          #Statistic
  "gtools",         #fold-change
  "RCy3",           #Cytoscape
  "glue",           #Saving plots 
  "colorBlindness", #Color palette
  "rScudo",         #Scudo
  "gProfileR",      #Interface to G-Profiler
  "gprofiler2",     #Interface to G-Profiler
  "BiocManager",
  "KEGGREST",
  "KEGGgraph",
  "AnnotationDbi",
  "org.Hs.eg.db",
  "pathfindR.data",
  "hgu95av2.db", #Genome annotation Affy HU 95
  "R.filesets",
  "hugene10sttranscriptcluster.db"
)

invisible(lapply(PACKAGES, library, character.only = TRUE))
```

```{r}
gse1<- getGEO("GSE53987") #GSE identifier of the file
gse<-gse1[[1]]
ex<- exprs(gse)# ex is the new data matrix
dim(ex) #54675 genes and 205 samples
meta<- pData(gse) #phenoData #save the meta data
```

```{r}
dataHippo<- ex[,1:68] 
dataBA46<- ex[,69:136] 
dataStriatum<- ex[,137:205] 
```


Normalization by log2 and housekeeping genes
```{r}
GADPH<-mean(log2(dataBA46["212581_x_at",]))
```

```{r}
dataBA46<- log2(dataBA46)
dataHippo<- log2(dataHippo)
dataStriatum<- log2(dataStriatum)
```

Redifine after the transposion
```{r}
dataBA46<- dataBA46/GADPH
dataHippo<- dataHippo/GADPH
dataStriatum<- dataStriatum/GADPH
```

```{r}
grpcol_BA46<- c(rep("#cc6a70ff", 17),rep("#253582ff", 19), rep("#7e4e90ff",17 ), rep("#f9a242ff",15))
Group_BA46 = c(rep("Bipolar", 17), rep("Control", 19), rep("Depression", 17), rep("Schizophrenia", 15))

grpcol_Hippo<- c(rep("#cc6a70ff", 18),rep("#253582ff", 18), rep("#7e4e90ff",17 ), rep("#f9a242ff",15))

Group_Hippo = c(rep("Bipolar", 18), rep("Control", 18), rep("Depression", 17), rep("Schizophrenia", 15))

grpcol_Striatum<- c(rep("#cc6a70ff", 17),rep("#253582ff", 18), rep("#7e4e90ff",16 ), rep("#f9a242ff",18))
Group_Striatum = c(rep("Bipolar", 17), rep("Control", 18), rep("Depression", 16), rep("Schizophrenia", 18))
```



Define the input
```{r}
pV <-0.001
gruppo1 <- "Schizophrenia"
gruppo2 <- "Depression"
i1 <- c(52:69)
i2 <- c(36:51)
data <- dataStriatum
group1_data <- data[,i1]
group2_data <- data[,i2]
```

```{r}
#assign a name to the dataframe
name <- paste(gruppo1, gruppo2, sep = "_vs_") 
assign(name, create_mine_data(data, gruppo1, gruppo2, group1_data, group2_data ))
```



```{r}
#ex_Striatum <-dataStriatum[c(result1, result2, result3, result4, result5, result6 ),] 
#dim(ex_Striatum) #we obtain 14083 significant genes
```



Let's transpose the matrix and add the Groups
```{r}
ex_Striatum_t <- t(ex_Striatum) #LDA need transposition
ex_Striatum_t <- cbind(as.data.frame(ex_Striatum_t),Group_Striatum) #LDA need labels, added to the data matrix
colnames(ex_Striatum_t)[ncol(ex_Striatum_t)] <- "Group"
duplicated_names <- duplicated(colnames(ex_Striatum_t))
dataStriatum<-ex_Striatum_t[!duplicated_names] #7490 after removing the duplicates
```

```{r}
save(dataStriatum, file = "dataStriatum.RData")
```



Separate the data into training and validation set
```{r}
set.seed(123)
# Sample 80% of the data for training and 20% for validation
train_indices <- sample(1:nrow(dataHippo), 0.8 * nrow(dataHippo))
train_data <- dataHippo[train_indices, ]
validation_data <- dataHippo[-train_indices, ]
```

Define the CV method
```{r}
metric <- "Accuracy"
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
```


Run the LDA
```{r}
library(MASS)
lda_model <- lda(Group ~ ., data = train_data)

# Predict on validation set
lda_predictions <- predict(lda_model, newdata = validation_data)$class

# Calculate accuracy
lda_accuracy <- mean(lda_predictions == validation_data$Group)
```
LDA ha 0.5 of accuracy for BA46, 0.4 for Striatum,  0.4 for Hippo


RF for feature selection
```{r}
set.seed(1)
fit.rf <- train(Group~., data = dataHippo, method="rf",
                metric=metric, trControl=control)

```

```{r}
confusionMatrix(fit.rf)
plot(fit.rf)
```
RF has accuracy BA46 0.625, Striatum 0.5





Since Rf is the one that better performed, let's extract the variable
```{r}
impp <- varImp(fit.rf)
impDF<-data.frame(impp$importance) 
geni_Hippo<-rownames(impDF)[order(impp$importance$Overall, decreasing = TRUE)]
geni_Hippo<- str_sub(geni_Hippo,2,-2)
write.csv(geni_Hippo[1:200], file = "geni_Hippo_200", quote=FALSE, row.names = FALSE)
#rownames(imp)
#rownames(impDF)[order(imp$Overall, decreasing=TRUE)]
#rownames(imp)[order(imp$Overall, decreasing=TRUE)]
```




Now from the converted list, remove the duplicats

```{r}
data <- readLines("geni_Striatum_200_ENTERGZ")
unique_data <- unique(data)
unique_data <- sort(unique_data)
write.table(unique_data, file = "geni_Striatum_200_ENTERGZ_unique.txt", quote=FALSE, row.names = FALSE, col.names = FALSE)
```


Upload the list already computed with LDA
```{r}
modified_gene_names <-readLines("geni_Striatum_200")
```

Now I want to filter the pairwise data in order to keep only the genes that are important for LDA
```{r}
filtered_Bipolar_vs_Control <- Bipolar_vs_Control[Bipolar_vs_Control $Gene_ID %in% modified_gene_names, ]

filtered_Bipolar_vs_Depression <- Bipolar_vs_Depression[Bipolar_vs_Depression$Gene_ID %in% modified_gene_names, ]

filtered_Bipolar_vs_Schizophrenia <- Bipolar_vs_Schizophrenia[Bipolar_vs_Schizophrenia$Gene_ID %in% modified_gene_names, ]

filtered_Control_vs_Depression <- Control_vs_Depression[Control_vs_Depression$Gene_ID %in% modified_gene_names, ]

filtered_Control_vs_Schizophrenia <- Control_vs_Schizophrenia [Control_vs_Schizophrenia$Gene_ID %in% modified_gene_names, ]

filtered_Schizophrenia_vs_Depression <- Schizophrenia_vs_Depression [Schizophrenia_vs_Depression$Gene_ID %in% modified_gene_names, ]
```

Save some results 
```{r}
save(filtered_Bipolar_vs_Schizophrenia, file = "filtered_Bipolar_vs_Schizophrenia.RData")
```


Identify the genes to their actual name

```{r}
x<- c(gse@featureData@data$ID)
y<- c(gse@featureData@data$`Gene Symbol`)
geni<- data.frame("Affy"=x,"Name"=y)
```

Upload the filtered datamatrix dataBA46

PHEAMAP
```{r}
df <- dataStriatum
#transpose and remove the group 
sig_expr_matrix <- t(df[, modified_gene_names]) #selct the 200 gene by LDA
geni_matrix <- geni$Name[geni$Affy %in% modified_gene_names]
rownames(sig_expr_matrix) <- geni_matrix
#change some gene name
```

```{r}
sample_annotations <- data.frame(Group_Striatum = Group_Striatum)
rownames(sample_annotations) <- colnames(sig_expr_matrix)
group_colors <- c("Bipolar" = "#cc6a70ff", "Control" = "#253582ff", 
                  "Depression" = "#7e4e90ff",  "Schizophrenia" = "#f9a242ff")
annotation_colors <- list(Group_Striatum = group_colors)
```


```{r}
pheatmap(sig_expr_matrix[c(1:11, 14, 17:20, 23:25, 28, 31, 35:37, 42:67), ], 
         scale = "row", 
         cluster_rows = TRUE, 
         annotation_col = sample_annotations,
         annotation_colors = annotation_colors,  # Custom group colors
         show_colnames = FALSE,
         cluster_cols = FALSE,
         main = "Heatmap of 50 Significant Genes Striatum",
         color = colorRampPalette(c("navy", "white", "red"))(50))
```




Volcano Plot

```{r}
df <- filtered_Control_vs_Schizophrenia
df$diffexpressed <- "NO"
# Categorize genes as "UP" if log2FoldChange > 0.6 and p-value < 0.05
df$diffexpressed[df$Fold_Change > 1 & df$P_Value < 0.01] <- "UP"
# Categorize genes as "DOWN" if log2FoldChange < -0.6 and p-value < 0.05
df$diffexpressed[df$Fold_Change < -1 & df$P_Value < 0.01] <- "DOWN"
df$padj <- p.adjust(df$P_Value, method = "fdr")

df$delabel <- NA
df$delabel[df$diffexpressed != "NO"] <-df$Gene_Symbol [df$diffexpressed != "NO"]
```


```{r}
mycolors <- c("#00AFBB", "#bb0c00", "grey")
names(mycolors) <- c("DOWN", "UP", "NO")

pCS<- ggplot(data=df, aes(x=Fold_Change, y=-log10(P_Value), col=diffexpressed, label=delabel)) + 

  geom_vline(xintercept=c(-1, 1), col= "grey", linetype = 'dashed') +
  geom_hline(yintercept=-log10(0.05), col= "grey", linetype = 'dashed')+
 scale_colour_manual(values = mycolors)+
 labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
 ggtitle('Volcano plot control vs schizophrenia  (Striatum)') +
  geom_point() + 
  theme_minimal()+
  geom_text()

```

```{r}

grid.arrange(pBC, pBD, pBS, pCD, pCS, pSD, nrow = 2, ncol = 3)
```

