---
title: "Untitled"
author: "Rayan Slatni"
date: "2023-07-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
gse<-loadRDS("GSE208338.rds")
```

```{r}
ex<- exprs(gse)# ex is the new data matrix
dim(ex) #54675 genes and 205 samples
meta<- pData(gse) #phenoData #save the meta data
```
```{r}
dataBA9<-ex
dataBA9<-dataBA9[, order(meta$`diagnosis:ch1`)]
```


Normalization step
```{r}
dataBA9<- log2(dataBA9)/GADPH
```

```{r}
grpcol_BA9<- c(rep("#cc6a70ff", 15),rep("#253582ff", 62), rep("#7e4e90ff",24 ), rep("#f9a242ff",68))
Group_BA9 = c(rep("Bipolar", 15), rep("Control", 62), rep("Depression", 24), rep("Schizophrenia", 68))
```

Define a Function to produce a Gene-pvalue-FC-group dataframe
```{r}
create_mine_data <- function(data, gruppo1, gruppo2, group1_data, group2_data){
  t_res <- t_test(data, gruppo1, gruppo2, i1, i2)
  result_with_fc <- data.frame(
    Exon <- rownames(data), #affymetrix name
    #compute the p value
    p_value <- t_res$p_value, 
    #compute the foldchange
    fold_change <- compute_fold_change(group1_data , group2_data), 
    Groups <- rep(paste(gruppo1, gruppo2, sep = "-"), nrow(data)))
  
   #filter for the sigificant genes
    filtered_results <- result_with_fc[t_res$significant_genes, ]
  #rename the columns
    colnames(filtered_results) <- c( "Gene_ID", "P_Value", "Fold_Change", "Comparison_Groups")
  return(filtered_results)
}
```


Define the input
```{r}
gruppo1 <- "Schizophrenia"
gruppo2 <- "Depression"
i1 <- c(102:169)
i2 <- c(78:101)
data <- dataBA9
group1_data <- data[,i1]
group2_data <- data[,i2]
```


```{r}
#assign a name to the dataframe
name <- paste(gruppo1, gruppo2, sep = "_vs_") 
assign(name, create_mine_data(data, gruppo1, gruppo2, group1_data, group2_data ))
```



```{r}
#ex_BA9 <- dataBA9[c(result1, result2, result3, result4, result5, result6 ),] 
#dim(ex_BA9) #we obtain 64332 significant genes
```


Let's transpose the matrix and add the Groups
```{r}
ex_BA9_t <- t(ex_BA9) #LDA need transposition
ex_BA9_t <- cbind(as.data.frame(ex_BA9_t),Group_BA9) #LDA need labels, added to the data matrix
colnames(ex_BA9_t)[ncol(ex_BA9_t)] <- "Group"
duplicated_names <- duplicated(colnames(ex_BA9_t))
dataBA9<-ex_BA9_t[!duplicated_names] #449 after removing the duplicates
```

```{r}
save(dataBA9, file = "dataBA9.RData")
```

Separate the data into training and validation set
```{r}
set.seed(123)

# Sample 80% of the data for training and 20% for validation
train_indices <- sample(1:nrow(dataBA9), 0.8 * nrow(dataBA9))
train_data <- dataBA9[train_indices, ]
validation_data <- dataBA9[-train_indices, ]
```


```{r}
library(MASS)
lda_model <- lda(Group ~ ., data = train_data)

# Predict on validation set
lda_predictions <- predict(lda_model, newdata = validation_data)$class

# Calculate accuracy
lda_accuracy <- mean(lda_predictions == validation_data$Group)
```
Accuracy of LDA 0.94


```{r}
mod.lda <- lda(Group ~ ., data=dataBA9)
mod.values <- predict(mod.lda, dataBA9)
```


Extract important variable for LDA
```{r}
lda_coefficients <- as.data.frame(mod.lda$scaling)
# Rank genes by their importance (absolute value of coefficients)
lda_coefficients$importance <- apply(lda_coefficients, 1, function(x) sum(abs(x)))
# Order genes by their importance
lda_coefficients <- lda_coefficients[order(-lda_coefficients$importance), ]
# Select the top 100 genes
top_200_genes <- head(lda_coefficients, 200)
# Extract the names of the top 100 genes
top_200_gene_names <- rownames(top_200_genes)
```

```{r}
remove_first_last_char <- function(x) {
  substr(x, 2, nchar(x) - 1)
}

# Apply the function to each gene name
gene_names <- sapply(top_200_gene_names, remove_first_last_char)
```

save the result
```{r}
write.table(gene_names, file = "geni_BA9_200.txt", quote=FALSE, row.names = FALSE, col.names = FALSE)
```

Upload the list already computed with LDA
```{r}
modified_gene_names <-readLines("geni_BA9_200_ENERGZ")
```

Merge them so we can have both names
```{r}
geni<- data.frame("Exon_id" = gene_names,"Name"= modified_gene_names)
```

Filter the datasets


```{r}
filtered_Bipolar_vs_Control <- Bipolar_vs_Control[Bipolar_vs_Control$Gene_ID %in% gene_names, ]

filtered_Bipolar_vs_Depression <- Bipolar_vs_Depression[Bipolar_vs_Depression$Gene_ID %in% gene_names, ]

filtered_Bipolar_vs_Schizophrenia <- Bipolar_vs_Schizophrenia[Bipolar_vs_Schizophrenia$Gene_ID %in% gene_names, ]

filtered_Control_vs_Depression <- Control_vs_Depression[Control_vs_Depression$Gene_ID %in% gene_names, ]

filtered_Control_vs_Schizophrenia <- Control_vs_Schizophrenia [Control_vs_Schizophrenia$Gene_ID %in% gene_names, ]

filtered_Schizophrenia_vs_Depression <- Schizophrenia_vs_Depression [Schizophrenia_vs_Depression$Gene_ID %in% gene_names, ]
```

Now add the normal gene name
```{r}
filtered_Control_vs_Schizophrenia$Gene_symbol <- geni$Name[gene_names %in%  filtered_Control_vs_Schizophrenia$Gene_ID]
```


Save some results

```{r}
save(filtered_Schizophrenia_vs_Depression, file = "filtered_Schizophrenia_vs_Depression.RData")
```


```{r}
df <- dataBA9
#transpose and remove the group 
sig_expr_matrix <- t(df[, gene_names]) #selct the 200 gene by LDA
geni_matrix <- geni$Name[geni$Exon_id %in% gene_names]
rownames(sig_expr_matrix) <- geni_matrix
```

```{r}
sample_annotations <- data.frame(Group_BA9 = Group_BA9)
rownames(sample_annotations) <- colnames(sig_expr_matrix)
group_colors <- c("Bipolar" = "#cc6a70ff", "Control" = "#253582ff", 
                  "Depression" = "#7e4e90ff",  "Schizophrenia" = "#f9a242ff")
annotation_colors <- list(Group_BA9 = group_colors)
```


```{r}
pheatmap(sig_expr_matrix[c(1:9, 11, 13:16,18:21, 23:25, 27:34, 36:42, 44:46, 48:58 ), ], 
         scale = "row", 
         cluster_rows = TRUE, 
         annotation_col = sample_annotations,
         annotation_colors = annotation_colors,  # Custom group colors
         show_colnames = FALSE,
         cluster_cols = FALSE,
         main = "Heatmap of 50 Significant Genes BA9",
         color = colorRampPalette(c("navy", "white", "red"))(50))
```

Volcano Plot

```{r}
df <- filtered_Control_vs_Schizophrenia
df$diffexpressed <- "NO"
# Categorize genes as "UP" if log2FoldChange > 0.6 and p-value < 0.05
df$diffexpressed[df$Fold_Change > 1 & df$P_Value < 0.01] <- "UP"
# Categorize genes as "DOWN" if log2FoldChange < -0.6 and p-value < 0.05
df$diffexpressed[df$Fold_Change < -1 & df$P_Value < 0.01] <- "DOWN"
df$padj <- p.adjust(df$P_Value, method = "fdr")

df$delabel <- NA
df$delabel[df$diffexpressed != "NO"] <-df$Gene_symbol [df$diffexpressed != "NO"]
```


```{r}
mycolors <- c("#00AFBB", "#bb0c00", "grey")
names(mycolors) <- c("DOWN", "UP", "NO")

pCS<- ggplot(data=df, aes(x=Fold_Change, y=-log10(P_Value), col=diffexpressed, label=delabel)) + 

  geom_vline(xintercept=c(-1, 1), col= "grey", linetype = 'dashed') +
  geom_hline(yintercept=-log10(0.05), col= "grey", linetype = 'dashed')+
 scale_colour_manual(values = mycolors)+
 labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) + 
 ggtitle('Volcano plot control vs schizophrenia (BA9)') +
  geom_point() + 
  theme_minimal()+
  geom_text()

```

```{r}
grid.arrange(pBC, pBD, pBS, pCD, pCS, pSD, nrow = 2, ncol = 3)
```



